---
title: Create Rmarkdown Document with SAS Code and Output - SAS engine
author: Melinda Higgins
date: '2019-10-01'
slug: create-rmarkdown-document-with-sas-code-and-output-sas-engine
categories:
  - R
  - SAS
  - Rmarkdown
tags:
  - R
  - SAS
  - Rmarkdown
subtitle: ''
summary: ''
authors: []
lastmod: '2020-09-02'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---


```{r, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

## Getting started - SAS Engine for Rmarkdown

To get started using SAS as your statistical software/data processing "engine" take a look at the following article: [http://www.ssc.wisc.edu/~hemken/SASworkshops/Markdown/SASmarkdown.html](http://www.ssc.wisc.edu/~hemken/SASworkshops/Markdown/SASmarkdown.html).

Also read up on the `SASmarkdown` package [https://cran.r-project.org/web/packages/SASmarkdown/](https://cran.r-project.org/web/packages/SASmarkdown/).

## Display the current `knitr` engine

The following Rmarkdown chunk shows the commands to see what are your current `knitr` engine settings.

Be sure to put `{r}` after the 3 backticks ``` to create the R chunk.

` ```{r} `

```
# see what the initial knitr engine settings are
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

` ``` `

```{r}
# see what the initial knitr engine settings are
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

## Setup the SAS engine

To get started you need:

1. Have SAS installed locally on your machine (i.e. you need a licensed copy)
2. You need to know where on your local drive that your SAS executable is located. Mine is located at `C:\Program Files\SASHome\SASFoundation\9.4\sas.exe`.
3. Install the `SASmarkdown` package [https://cran.r-project.org/web/packages/SASmarkdown/index.html](https://cran.r-project.org/web/packages/SASmarkdown/index.html).
4. Then setup your `knitr` options as follows (type these commands into an R chunk - see next section below):

```
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -linesize 75"
knitr::opts_chunk$set(engine="sashtml", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)
```

## Change settings to use HTML output from SAS

NOTE: You will need to install the `SASmarkdown` package first. For this exercise we're using `engine="sashtml"` which works well when knitting to HTML. However, this format will not work if you knit to PDF or DOC. This leverages the ODS output from SAS.

The R chunk below loads the `SASmarkdown` package, changes the `knitr` engine to `"sashtml"` and runs the `knitr::opts_chunk$get()` commands again to make sure that the engine has been switched from `"R"` to `"sashtml"`.

` ```{r} `

```
# load the SASmarkdown package
library(SASmarkdown)

# set up the options so that knit knows where you SAS executable is
# set the linesize to be easily readable on letter size paper, portrait
# and set the knir options using opts_chunk$set().
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -linesize 75"
knitr::opts_chunk$set(engine="sashtml", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)

# run these commands to convince yourself that
# within this knitr session the engine changed.
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

` ``` `

```{r}
# load the SASmarkdown package
library(SASmarkdown)

# set up the options so that knit knows where you SAS executable is
# set the linesize to be easily readable on letter size paper, portrait
# and set the knir options using opts_chunk$set().
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -linesize 75"
knitr::opts_chunk$set(engine="sashtml", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)

# run these commands to convince yourself that
# within this knitr session the engine changed.
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

## If you want to use SAS but make PDF or DOC files

If you want to knit to PDF or DOC, you should change the engine to `engine="sas"`. This will NOT use ODS output from SAS.

These R chunk commands are provided for reference but not executed here.

```
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -linesize 75"
knitr::opts_chunk$set(engine="sas", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)
```

## Try some `SAS` code

This code chunk below runs the `PROC MEANS` command from SAS using the built in dataset `sashelp.class`.

` ```{r} `

```
proc means data=sashelp.class;
run;
```

` ``` `

```{r}
proc means data=sashelp.class;
run;
```

## More SAS code 

Here is another example of SAS code chunk using the `PROC CORR` commands to see correlations between the variables in the `sashelp.class` dataset and  visualize the scatterplot matrix.

` ```{r} `

```
proc corr data=sashelp.class plots=matrix;
run;
```

` ``` `

```{r}
proc corr data=sashelp.class plots=matrix;
run;
```

## Reset engine to R

While this is great using SAS, if you want to switch back to using R within the same Rmarkdown document, within the same `knitr` session, you'll need to tell `knitr` that you are switching engines.

To get this next chunk to run, you'll need to reset the `knitr` engine within the chunk options directly and then use the command `knitr::opts_chunk$set(engine="R", engine.path=NULL, engine.opts=NULL, comment=NA)` to reset back to R.

Be sure to put `{r, engine='R'}` after the 3 backticks ``` to set this R chunk back to the R engine.

` ```{r, engine='R'} `

```
# this chunk has the engine set back to `R`

# run a short bit of r code
# scatterplot of cars dataset
plot(cars)

# check the current engine
knitr::opts_chunk$get()$engine

# reset the engine globally (i.e. so it will work outside of
# this chunk so you don't have to keep typing engine=`R`)
knitr::opts_chunk$set(engine="R",  engine.path=NULL, 
        engine.opts=NULL, comment=NA)

# confirm that this change was applied
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

` ``` `

```{r, engine='R'}
# this chunk has the engine set back to `R`

# run a short bit of r code
# scatterplot of cars dataset
plot(cars)

# check the current engine
knitr::opts_chunk$get()$engine

# reset the engine globally (i.e. so it will work outside of
# this chunk so you don't have to keep typing engine=`R`)
knitr::opts_chunk$set(engine="R",  engine.path=NULL, 
        engine.opts=NULL, comment=NA)

# confirm that this change was applied
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

...and here is another R chunk with no engine defined but the engine was switched back to R in the previous chunk.

` ```{r} `

```
# this chunk didn't list the engine explicitly, so let's
# make sure the global change carried over into this chunk
knitr::opts_chunk$get()$engine

# and some more R code
# simple linear regression of distance by speed
# for cars dataset
summary(lm(dist ~ speed, data = cars))
```

` ``` `

```{r}
# this chunk didn't list the engine explicitly, so let's
# make sure the global change carried over into this chunk
knitr::opts_chunk$get()$engine

# and some more R code
# simple linear regression of distance by speed
# for cars dataset
summary(lm(dist ~ speed, data = cars))
```

Good luck creating documents with Rmarkdown for SAS code and output.

#### Read more on R and SAS

* [R-Bloggers](https://www.r-bloggers.com/) 
* [PROC-X SAS Bloggers](http://www.proc-x.com/)
